Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-07-27 TEX/draft_biology.tex
method section

\paragraph{Aperiodic autocorrelation functions.}
The second-order autocorrelation of  an arbitrary image $z\in\mathbb{R}^{m\times m}$ is defined by
  \begin{align} 
 a_z^2[\Bell] & = \frac{1}{m^2} \sum_{i_1 = \max\{0, -\ell_1\}}^{m-1 + \min\{0, -\ell_1\}} \sum_{i_2 = \max\{0, -\ell_2\}}^{m-1 + \min\{0, -\ell_2\}}z[\ii]z[\ii+\Bell]. 
 \end{align}
 Below, the image $z$ might be either a target image $x$ or the micrograph $y$. 
 It can be shown that in the limit of infinitely many micrographs $K\to\infty$,  the mean pixel value of the micrographs is equal to the mean pixel value of the signal $x$, scaled by  
 \begin{align}
 \gamma & = \frac{M P^2}{N^2},
 \end{align}
 where $M$ is the number of occurrences of $x$ across micrographs. 
 This  is the fraction of entries of all micrographs occupied by occurrences of $x$, that is, the  density of the signal. 
  The spacing constraint~\eqref{eq:spacing} imposes $\gamma\leq\frac{P}{2P-1}\approx 1/2$.

 % Similarly, as $K\to\infty$ the second-order autocorrelations are related through 
 Similarly, numbering micrographs as $y_1, \ldots, y_K$, the empirical average of the second-order autocorrelations of the micrographs converges as follows:
 \begin{align} \label{eq:a2}
	\lim_{K \to \infty} \frac{1}{K} \sum_{k=1}^{K} a_{y_k}^2 & = \gamma a_{x}^2 + \sigma^2\delta,
 \end{align}
 where $\delta[\Bell]=1$ for $\Bell=(0,0)$ and zero otherwise.
 Thus, for any noise level and without the need to even implicitly locate the images $x$ in the micrographs, given sufficiently many micrographs we are able to compute the second-order autocorrelation of $x$ itself. Recovering $x$ from $a_x^2$ can be done up to symmetries, as detailed below.
 Equation~\eqref{eq:a2} is proven and its extension to higher-order autocorrelations and multiple distinct planted images is discussed in~\cite{bendory2018estimation}.

\paragraph{Numerical experiment with a 2-D image.}

@@

\subsection{3-D experiment}
\TODO{Add similarities to 2-D}

\TODO{Proposal: Use W for size of projection, M for size of micrograph}

@@

Let $\II\in\RNN$ denote a micrograph. We assume that the micrograph consists of shifted copies of projections contaminated by additive white Gaussian noise:
\begin{equation}\label{eq:micrograph_model}
\II = \sum_{t=1}^{M} I_{\omega_t}\ast\delta_{\mb s_t}+ \varepsilon, \quad \varepsilon~\sim\mathcal{N}(0,\sigma^2 I),
\end{equation}
where the viewing directions $\omega_t$ are assumed to be drawn from the uniform distribution over $\text{SO}(3)$, $\mb s_t$ denotes the location of the center of the $t$th projection in the micrograph. Similarly to~\eqref{eq:spacing}, we impose a separation condition so that any two projections are separated by at least $2L-1$ \TODO{2L?} pixels in each direction.
%similarly to~\eqref{eq:spacing}:
%\be\label{eq:sep_cond} \ ||\mb s_j - \mb s_i||_{\infty} \geq 2L. \ee
Note that~\eqref{eq:model} can be also written as a sum of Delta functions as in~\eqref{eq:micrograph_model} 
%\TODO{Say something about nonuniformity in practice? Remember that biologists will jump at this...}

Define the $p$th autocorrelation of $\II$ as
\begin{equation*} \label{eq:Kth_autocorrelation}
a^p_\II(\Bell_1,\ldots, \Delta \Bell_{p-1}) = \frac{1}{N^2}\sum_{i,j = 1}^M\II(\mb i)\II(\mb i+\Bell_1)\cdots \II(\mb i+\Bell_{p-1}),
\end{equation*}
where $\mb i = (i,j)$. %, $\Delta \mb i = (\Delta i, \Delta j)$.% and we set $\II(i,j)=0$ if either $i,j\notin\{1,\ldots, M\}$. 
%In our procedure, we compute the first three autocorrelations. %, called the mean, power spectrum, and bispectrum, of the micrographs. \TODO{This definition and terminology should come earlier}. 
Under the specified condition, we show in Appendix TKTK that the first three autocorrelations  of the micrograph are related to those of the projections by 
\begin{align} 
\lim_{M\to\infty}a^p_\II  = \gamma\langle a^k_{I_\omega}\rangle_{\omega} + b_p, \quad p=1,2,3,
\end{align}
where $\langle\cdot\rangle$ denotes averaging and $b_k$ is a bias term. %and we take the limit as the number of projections $N$, and hence the dimensions of the micrograph $M$, tend to $\infty$. 
More precisely,  $b_1 = 0$ and therefore the mean is unbiased. The bias term of the second-order autocorrelation  $b_2$ depending only on $\sigma^2$, the variance of the noise. Hence, if the noise level can be accurately estimated from the micrograph, this bias can be removed. 
The bias term of the third-order autocorrelation $b_3$ depends on the mean of the micrograph and $\sigma^2$.  Therefore, given enough projections, we can accurately estimate the quantities $\gamma\langle m_k[I_{\omega}]\rangle_{\omega}$ directly from the micrograph.

Since we assume that the viewing directions $\omega_j$ are uniformly distributed over $\text{SO}(3)$, we would like to average over all in-plane rotations of the projections in the computation of the autocorrelations. To do so, we follow~\cite{landa2017steerable,zhao2016fast} and expand the second and third autocorrelations of the micrograph in Prolate Spheroidal Wave Functions (PSWFs). 
It can then be shown that the expansion coefficients $\mathfrak{m}_k$ of the autocorrelations in PSWFs are related to the expansion coefficients of the volume by

\begin{align}\label{eq:coeffs_to_moms}
&\mathfrak{m}_1\{a_{\ell,m,s}\} = \frac{L}{\sqrt{4\pi}}\sum_sa_{0,0,s}j_{0,s}(0) \\ 
&\mathfrak{m}_2\{a_{\ell,m,s}\}(q) = \frac{\sqrt{2\pi}}{4\pi}\sum_{\substack{\ell,m\\s_1,s_2}}C_2^{(q)}(\ell,s_1,s_2)a_{\ell,m,s_1}\overline{a_{\ell,m,s_2}}
 \\
&\mathfrak{m}_3\{a_{\ell,m,s}\}(k,q_1,q_2) = \sum_{\substack{\ell_1,m_1,s_1\\\ell_2,m_2,s_2\\s_3}}\sum_{\ell_3=|\ell_1-\ell_2|}^{\min(L,\ell_1+\ell_2)}C_3^{(k,q_1,q_2)}(\ell_1,m_1,s_1;\ell_2,m_2,s_2;\ell_3,s_3)a_{\ell_1,m_1,s_1}a_{\ell_2,m_2,s_2}\overline{a_{\ell_3,m_1+m_2,s_3}},
\end{align}
see Appendix TKTK for the definitions of $C_2,C_3$.

To recover the expansion coefficients $\{a_{\ell,m,s}\}$ from the autocorrelation coefficients estimated from the micrograph $\mk m_k[\II]$, we minimize the weighted least-squares problem
\be\label{eq:min_problem_cryo} \min_{\{a_{\ell,m,s}\}} \sum_{k=1}^3\frac{\Big|\Big|\mk m_k\{a_{\ell,m,s}\} - \mk m_k[\II]\Big|\Big|_F^2}{||\mk m_k[\II]||_F^2}.\ee

>>>>

%\paragraph{Aperiodic autocorrelation functions.}
The second-order autocorrelation of  an arbitrary image $z\in\mathbb{R}^{m\times m}$ is defined by
  \begin{align} 
 a_z^2[\Bell] & = \frac{1}{m^2} \sum_{i_1 = \max\{0, -\ell_1\}}^{m-1 + \min\{0, -\ell_1\}} \sum_{i_2 = \max\{0, -\ell_2\}}^{m-1 + \min\{0, -\ell_2\}}z[\ii]z[\ii+\Bell]. 
 \end{align}
 Below, the image $z$ might be either a target image $x$ or the micrograph $y$. 
 It can be shown that in the limit of infinitely many micrographs $K\to\infty$,  the mean pixel value of the micrographs is equal to the mean pixel value of the signal $x$, scaled by  
 \begin{align}
 \gamma & = \frac{M P^2}{N^2},
 \end{align}
 where $M$ is the number of occurrences of $x$ across micrographs. 
 This  is the fraction of entries of all micrographs occupied by occurrences of $x$, that is, the  density of the signal. 
  The spacing constraint~\eqref{eq:spacing} imposes $\gamma\leq\frac{P}{2P-1}\approx 1/2$.

 % Similarly, as $K\to\infty$ the second-order autocorrelations are related through 
 Similarly, numbering micrographs as $y_1, \ldots, y_K$, the empirical average of the second-order autocorrelations of the micrographs converges as follows:
 \begin{align} \label{eq:a2}
	\lim_{K \to \infty} \frac{1}{K} \sum_{k=1}^{K} a_{y_k}^2[\Bell] & = \gamma a_{x}^2[\Bell] + \sigma^2\delta_{(0,0)}[\Bell], \quad \Bell\in[0,P-1]^2,
 \end{align}
 where $\delta_{(i,j)}[\Bell]=1$ for $\Bell=(i,j)$ and zero otherwise.
 Thus, for any noise level and without the need to even implicitly locate the images $x$ in the micrographs, given sufficiently many micrographs we are able to compute the second-order autocorrelation of $x$ itself. Recovering $x$ from $a_x^2$ can be done up to symmetries, as detailed below.
 Equation~\eqref{eq:a2} is proven and its extension to higher-order autocorrelations and multiple distinct planted images is discussed in~\cite{bendory2018estimation}.

%\paragraph{Numerical experiment with a 2-D image.}

@@

\subsection{3-D experiment}

@@

Let $\II\in\RNN$ denote a micrograph. We assume it consists of shifted copies of projections contaminated by additive white Gaussian noise:
\begin{equation}\label{eq:micrograph_model}
\II = \sum_{t=1}^{M} I_{\omega_t}\ast\delta_{\mb s_t}+ \varepsilon, \quad \varepsilon~\sim\mathcal{N}(0,\sigma^2 I),
\end{equation}
where the viewing directions $\omega_t$ are assumed to be drawn from the uniform distribution over $\text{SO}(3)$ and $\mb s_t$ denotes the location of the center of the $t$th projection in the micrograph. Similarly to~\eqref{eq:spacing}, we impose a separation condition so that any two projections are separated by at least $2L-1$ \TODO{2L?} pixels in each direction.
%similarly to~\eqref{eq:spacing}:
%\be\label{eq:sep_cond} \ ||\mb s_j - \mb s_i||_{\infty} \geq 2L. \ee
Note that~\eqref{eq:model} can be also written as a sum of $\delta$ functions as in~\eqref{eq:micrograph_model}. 
%\TODO{Say something about nonuniformity in practice? Remember that biologists will jump at this...}

Define the $p$th autocorrelation of $\II$ as
\begin{equation*} \label{eq:Kth_autocorrelation}
a^p_\II(\Bell_1,\ldots, \Bell_{p-1}) = \frac{1}{N^2}\sum_{\mb i}\II(\mb i)\II(\mb i+\Bell_1)\cdots \II(\mb i+\Bell_{p-1}),
\end{equation*}
where the summation is for $\mb i $ ranging over the $N^2$  pixels of the micrograph. %, $\Delta \mb i = (\Delta i, \Delta j)$.% and we set $\II(i,j)=0$ if either $i,j\notin\{1,\ldots, M\}$. 
%In our procedure, we compute the first three autocorrelations. %, called the mean, power spectrum, and bispectrum, of the micrographs. \TODO{This definition and terminology should come earlier}. 
Let $\II_1,\ldots\II_K$ denote a set of $K$ micrographs. 
Under the specified conditions, we show in Appendix TKTK that the first three autocorrelations  of the micrographs are related to those of the projections by \TODO{To fix}
\begin{align} \label{eq:ac_micrographs}
\lim_{K\to\infty}\frac{1}{K}\sum_{i=1}^Ka^p_{\II_i}[\Bell_1,\ldots,\Bell_{p-1}]  &= \gamma\left\langle a^p_{I_\omega}[\Bell_1,\ldots,\Bell_{p-1}]\right\rangle_{\omega} + b_p[\Bell_1,\ldots,\Bell_{p-1}], \\ p=1,2,3,& \quad \Bell_1,\ldots,\Bell_{p-1}\in[0,P-1]^2, \nonumber
\end{align}
where $\langle\cdot\rangle_\omega$ denotes averaging over all possible viewing directions $\omega$ and $b_p$ is a bias term. %and we take the limit as the number of projections $N$, and hence the dimensions of the micrograph $M$, tend to $\infty$. 
Specifically,  $b_1 = 0$ and therefore the mean is unbiased. The bias term of the second-order autocorrelation  $b_2$ depends only on $\sigma^2$, the variance of the noise. Hence, if the noise level can be accurately estimated from the micrographs, this bias can be removed. 
Finally, the bias term of the third-order autocorrelation $b_3$ depends on the mean of the micrograph and $\sigma^2$.  Therefore, given sufficiently many projections, we can accurately estimate the quantities $\gamma\langle a^p_{I_{\omega}}\rangle_{\omega}$ directly from the micrographs. These quantities are functions of the unknown coefficients $\tamir_{l,m,s}$ and we could proceed to invert their relation, as we did in the 2-D toy example.

In practice, we want to leverage one more feature of the 3-D reconstruction problem.
Since all in-plane rotations of the micrographs are equally likely observations, 
it is desirable in~\eqref{eq:ac_micrographs} to average over all in-plane rotations as well.
This can be done efficiently using Prolate Spheroidal Wave Functions (PSWFs). 
The details are in the appendix.

% as follows. We scan the micrographs with a $P\times P$ window. We expand each window in Prolate Spheroidal Wave Functions (PSWFs) 
%and compute the  second and third autocorrelations in that basis. In PSWFs, it is straightforward to average over all in-plane rotations~\cite{landa2017steerable,zhao2016fast}. We further average over all windows, resulting in autocorrelation in the PSWF coefficients.
%It can be shown that the PSWF coefficients of the data $\mathfrak{a}^p$  are related to the expansion coefficients of the volume~\eqref{eq:volume_expansion} by
%\begin{align}\label{eq:coeffs_to_moms}
%&\mathfrak{a}^1_x = \frac{L}{\sqrt{4\pi}}\sum_sa_{0,0,s}j_{0,s}(0), \\ 
%&\mathfrak{a}^2_x(q) = \frac{1}{2\sqrt{2\pi}}\sum_{\substack{\ell,m\\s_1,s_2}}C_2^{(q)}(\ell,s_1,s_2)a_{\ell,m,s_1}\overline{a_{\ell,m,s_2}},
% \\
%&\mathfrak{a}^3_x(k,q_1,q_2) = \sum_{\substack{\ell_1,m_1,s_1\\\ell_2,m_2,s_2\\s_3}}\sum_{\ell_3=|\ell_1-\ell_2|}^{\min(L,\ell_1+\ell_2)}C_3^{(k,q_1,q_2)}(\ell_1,m_1,s_1;\ell_2,m_2,s_2;\ell_3,s_3)a_{\ell_1,m_1,s_1}a_{\ell_2,m_2,s_2}\overline{a_{\ell_3,m_1+m_2,s_3}}.
%\end{align}
%The definitions of $C_2$ and $C_3$ are provided in Appendix TKTK.
%
%To recover the expansion coefficients $\{a_{\ell,m,s}\}$ from the autocorrelation coefficients estimated from the micrograph $\mk m_k[\II]$, we minimize the weighted least-squares problem
%\be\label{eq:min_problem_cryo} \min_{\{a_{\ell,m,s}\}} \sum_{k=1}^3\frac{\Big|\Big|\mk m_k\{a_{\ell,m,s}\} - \mk m_k[\II]\Big|\Big|_F^2}{||\mk m_k[\II]||_F^2}.\ee
